---
title: "Dplyr and Writing Scripts"
author: "Yanru Xing"
---

*Adapted from [Data Science for R](r4ds.had.co.nz) by Hadley Wichkam and Garrett Grolemund*

Download slides [here](Powerpoints/dplyr.pdf)

## tidyverse = A collection of R packages  
\  


### Package ‘dplyr’  
#### A Grammar of Data Manipulation  
#### Description A fast, consistent tool for working with data frame like objects, both in memory and out of memory.  



```{r dplyr, echo=T}
#install.packages("nycflights13")
#install.packages("tidyverse")

library(tidyverse)
library(nycflights13)

nycflights13::flights #This data frame contains all 336,776 flights that departed from New York City in 2013
flights
#View(flights) #To see the whole dataset, which will open the dataset in the RStudio viewer.
```
#### Five key dplyr functions that allow you to solve the vast majority of your data manipulation challenges:
* Pick observations by their values (filter()).
* Reorder the rows (arrange()).
* Pick variables by their names (select()).
* Create new variables with functions of existing variables (mutate()).
* Collapse many values down to a single summary (summarise()).

##### Filter Rows with filter()
###### filter() allows you to subset observations based on their values. The first argument is the name of the data frame. The second and subsequent arguments are the expressions that filter the data frame. For example, we can select all flights on January 1st with:

```{r Comparisons, echo=T}
filter(flights, month == 1, day == 1)
jan1 <- filter(flights, month == 1, day == 1)
jan1
# (dec25 <- filter(flights, month == 12, day == 25)) #R either prints out the results, or saves them to a variable. If you want to do both, you can wrap the assignment in parentheses
```

#### Comparison Operators

```{r Comparisons_operatorx, echo=T}
# There’s another common problem you might encounter when using ==: floating point numbers. These results might surprise you!

sqrt(2) ^ 2 == 2
1 / 49 * 49 == 1

#Computers use finite precision arithmetic (they obviously can’t store an infinite number of digits!) so remember that every number you see is an approximation. Instead of relying on ==, use near():
near(sqrt(2) ^ 2,  2)
near(1 / 49 * 49, 1)

```

#### Logical Operators 

```{r Logical_operators, echo=T}
# The following code finds all flights that departed in November or December:
filter(flights, month == 11 | month == 12)
nov_dec <- filter(flights, month %in% c(11, 12))
nov_dec
# Find flights that weren’t delayed (on arrival or departure) by more than two hours, you could use either of the following two filters:
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```
#### One important feature of R that can make comparison tricky are missing values, or NAs (“not availables”). NA represents an unknown value so missing values are “contagious”: almost any operation involving an unknown value will also be unknown.
```{r Missing_Value, echo=T}
NA > 5
10 == NA
NA + 10
NA / 2
NA == NA
x <- NA
y <- NA
x == y
is.na(x) # Function to determine is a value is missing

```
# filter() & missing values

```{r filter_missing, echo=T}
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)

filter(df, is.na(x) | x > 1)

```
#### filter() exercises
```{r filter_exercises, echo=T}
# 1. Find all flights that:
#    1. Had an arrival delay of two or more hours
filter(flights, (arr_delay > 120))

#    2. Flew to Houston (IAH or HOU)
filter(flights, (dest == "IAH" | dest == "HOU"))

#    3. Were operated by United, American, or Delta
filter(flights, (carrier %in% c("UA","AA","DL")))

#    4. Departed in summer (July, August, and September)
filter(flights, (month == 7 | month == 8 | month == 9 ))

#    5. Arrived more than two hours late, but didn’t leave late
filter(flights, (arr_delay > 120 & dep_delay <= 0))

#    6. Were delayed by at least an hour, but made up over 30 minutes in flight


#    7. Departed between midnight and 6am (inclusive)
filter(flights, (dep_time >= 1 & dep_time <= 600))   

# 2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?

# 3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

# 4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)


```
#### Arrange rows with arrange()
##### arrange() works similarly to filter() except that instead of selecting rows, it changes their order. It takes a data frame and a set of column names (or more complicated expressions) to order by. 
```{r arrange, echo=T}
# If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:
arrange(flights, year, month, day)

# Use desc() to re-order by a column in descending order:
arrange(flights, desc(dep_delay))
```
# Missing values are always sorted at the end:
```{r arrange1, echo=T}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)

arrange(df, desc(x))
```
#### Exercises

##### 1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

##### 2. Sort flights to find the most delayed flights. Find the flights that left earliest.

##### 3. Sort flights to find the fastest flights.

##### 4. Which flights travelled the longest? Which travelled the shortest?

#### Select Columns with select() 
##### It’s not uncommon to get datasets with hundreds or even thousands of variables. In this case, the first challenge is often narrowing in on the variables you’re actually interested in. 

##### select() allows you to rapidly zoom in on a useful subset using operations based on the names of the variables.

#### select()
```{r select, echo=T}
# Select columns by name
select(flights, year, month, day)
# Select all columns between year and day (inclusive)
select(flights, year:day)
# Select all columns except those from year to day (inclusive)
select(flights, -(year:day))
```
##### There are a number of helper functions you can use within select():

* starts_with("abc"): matches names that begin with “abc”.

* ends_with("xyz"): matches names that end with “xyz”.

* contains("ijk"): matches names that contain “ijk”.

```{r select1, echo=T}
select(flights, starts_with("dep"))
select(flights, ends_with("lay"))
select(flights, contains("ime"))

```

#### Add new variables with mutate()
##### mutate() adds new columns at the end of your dataset, new columns that are functions of existing columns

```{r mutate, echo=T}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
# Note that you can refer to columns that you’ve just created:
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)

# If you only want to keep the new variables, use transmute():
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```
#### Grouped summaries with summarise()
```{r summarise, echo=T}
# summarise() collapses a data frame to a single row:
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
#summarise() is not terribly useful unless we pair it with group_by(). This changes the unit of analysis from the complete dataset to individual groups. Then, when you use the dplyr verbs on a grouped data frame they’ll be automatically applied “by group”. For example, if we applied exactly the same code to a data frame grouped by date, we get the average delay per date:
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```
#### Combining multiple operations with the pipe

##### Imagine that we want to explore the relationship between the distance and average delay for each location. Using what you know about dplyr, you might write code like this:

```{r pipe, echo=T}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
delay <- filter(delay, count > 20, dest != "HNL")
# It looks like delays increase with distance up to ~750 miles 
# and then decrease. Maybe as flights get longer there's more 
# ability to make up delays in the air?
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)

```
##### Where missing values represent cancelled flights, we could also tackle the problem by first removing the cancelled flights. 
```{r pipe1, echo=T}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))

```

##### Grouping is most useful in conjunction with summarise(), but you can also do convenient operations with mutate() and filter():
```{r Grouped_mutates, echo=T}
# Find the worst members of each group:
flights_sml %>% 
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)

# Find all groups bigger than a threshold:
popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 365)
popular_dests

# Standardise to compute per group metrics:
popular_dests %>% 
  filter(arr_delay > 0) %>% 
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>% 
  select(year:day, dest, arr_delay, prop_delay)


```

