---
title: "Introduction to R"
---

### Set working directory

First, set your working directory. This is a folder in which you will save this script, along with any files that will be used as part of your analyses. 



### R is essentially a fancy calculator:
```{r calculator}
1+1
3*2
8^8
```


To make programming simpler, we usually save values or sets of values that we want to use again as "objects". We do this using the symbols "<-" (called the "gets arrow"). 
```{r}
a <- 2 + 2
a
b <- a + 2
b
```



Comparing values will return "True" or "False"
```{r,results='hide'}
2==2 ## does equal 
2!=2 ## does not equal
```


>#### Note: R is case-sensitive and unforgiving! 



##Getting Started 
First, lets do some fun stuff! To get started, install the *package* "ggplot2". Packages are toolboxes that include functions, data and code for specific tasks. Note, remove the "#" from before install.packages() to load for the first time. The "#" comments out any code so it will not be run.

```{r}
#install.packages("ggplot2")
library(ggplot2)
```

> #### Note: if you have installed an R package once on your computer, you will not need to install it again. You will need to *load* the library at the begining of each new session using the library() function




<!-- ## Getting Data Into R -->

<!-- Now make sure to place your "StripeRust.csv" file in the working directory that we set, then load the data in using the "read.csv" *function*.  -->
<!-- ```{r} -->
<!-- RustData <- read.csv("StripeRust.csv", header = TRUE)  -->
<!--   #Header = TRUE lets R know that we have headings in the first row of our data set.  -->

<!-- head(RustData) ## check the first few rows of the dataset -->
<!-- names(RustData) ## check the column headers -->
<!-- ``` -->


<!-- > #### Note: this data is disease severity ratings taken in a wheat stripe rust nursury over the course of two seasons. "DAI" = days after inoculation.  -->


<!-- We can plot this data using ggplot2. Tell the graphing function to use the data "RustData" and then add a layer to the plot with the geom_point() function with severity on the y axis and time on the x axis. Color the points by the year that data was collected.  -->
<!-- ```{r} -->
<!-- ggplot(data = RustData) + -->
<!--   geom_point(mapping = aes(x = DAI, y = Severity, color = Year)) -->
<!-- ``` -->

<!-- >#### Note: It looks like disease took longer to develop in year 2, and was less severe by day 100. -->


<!-- If we switch from the function "geom_point" to "geom_line" we can plot lines and not points.   -->
<!-- ```{r} -->
<!-- ggplot(data = RustData) + -->
<!--   geom_line(mapping = aes(x = DAI, y = Severity, color = Year)) -->
<!-- ``` -->

<!-- We can look at the graphs side-by-side using the facet_wrap() function.  -->
<!-- ```{r} -->
<!-- ggplot(data = RustData) + -->
<!--   geom_point(mapping = aes(x = DAI, y = Severity)) + -->
<!--   facet_wrap(~Year)  -->
<!-- ``` -->

<!-- > #### Note: be sure to include a "+" at the end of each line, until you want to close the command.    -->

<!-- Finally, lets add a title!  -->
<!-- ```{r} -->
<!-- ggplot(data = RustData) + -->
<!--   geom_point(mapping = aes(x = DAI, y = Severity, color = Year)) + -->
<!--   facet_wrap(~Year) + -->
<!--   labs(title = "Stripe Rust Severity Over Time") -->
<!-- ``` -->

## Vectors

> ####Vectors are a way to set a series of data elements as an object. 

```{r}
v1 <- c(1, 2, 3, 4, 5) #numbers
v2 <- c("hello", "world") #characters 
v3 <- c("TRUE", "FALSE", "TRUE") #logical values (also could be - "T", "F", "T")
```


Lets make a vector with hypothetical ratings of "R expertise" on a scale of 1-10.  
```{r}
WithR <- c(8.5, 6.5, 4, 1, 3, 10, 5, 5, 5, 1, 1, 6, 6)
WithR
```

We can use the following functions to look at some summary statistics. 
```{r}
summary(WithR)
mean(WithR)
sd(WithR)
```

And graph a histogram of this distribution!
```{r}
hist(WithR, xlab = "Self-Reported R Proficiency")
```


> ###Notice that I choose to name my new object "WithR". 
> ###It is important to note that:
  - R objects are case sentitive so "WithR" is different than "Withr" or "withr"
  - Naming of objects is personal preference but the more intuitive, the better 
  - You cannot start with a number, but can end with one (WithR2)
  - Object names can only contain letters, number, "_", and "." (WithR_2 or WithR.2)
  
When reading the assignment statement below we read it as "ClassSize gets 13"
```{r}
ClassSize <- 13 # Assingment
ClassSize
ClassSize <- length(WithR) ## legnth() will also give us the size of WithR
ClassSize
```

Lets make a vector of names. 
```{r}
Names <- c("Jim", "Carole", "Joe", "Michelle", "Jen", "Pete", "Paul", "Tim", 
           "Jess", "Mark", "Jill", "Cam", "Kate") ## names are random to preserve anonymity
Names
```

We can bind together two or more vectors into a dataframe using cbind.data.frame() to bind columns. Note that cbind stands for "column bind" and rbind stands for "row bind".
```{r}
NamesScores <- cbind.data.frame(Names, WithR) #column bind
NamesScores
```


Lets bind vertically!
```{r}
NamesScoresR <- rbind(Names, WithR) # row bind
NamesScoresR <- as.data.frame(NamesScoresR) # make a dataframe
NamesScoresR
```


If we want to add the age and gender of the students, we can add a vector of names as another column to the data set

```{r}
Age <- c(22, 31, 25, 21, 22, 35, 42, 27, 26, 33, 26, 28, 22) #make a vector of ages
Gender <- c("Male", "Female", "Male", "Female", "Female", "Male","Male","Male", "Female", "Male", "Female", "Male", "Female")
NamesScores <- cbind(NamesScores, Age, Gender) #column bind
NamesScores
```


Lets save this new dataframe as a csv in our working directory. Check to see if you have the file in your folder!

```{r}
write.csv(NamesScores, "NamesScores.csv")
```

Alternatively, we could have made our data frame like this: 
```{r}
NamesScores3 <- data.frame(
                    Names =c("Jim", "Carole", "Joe", "Michelle", "Jen", "Pete", "Paul", "Tim", 
                            "Jess", "Mark", "Jill", "Cam", "Kate") ,
                    Experience = c(8.5, 6.5, 4, 1, 3, 10, 5, 5, 5, 1, 1, 6, 6) , 
                    Age = c(22, 31, 25, 21, 22, 35, 42, 27, 26, 33, 26, 28, 22) , 
                    Gender = c("Male", "Female", "Male", "Female", "Female", "Male","Male","Male", "Female", "Male", "Female", "Male", "Female"))
NamesScores3            
```



###Subsetting

What if we only care about the first four members of the class? We can use *subsetting* to select the values in our data frame that we are interested in. 
```{r}
## syntax: dataframe[rows, columns]
FirstFour <- NamesScores[1:4,] ## this means select row 1 through 4 in data frame NamesScores
FirstFour
```

What if we just wanted the second column?
```{r}
SecondColumn <- NamesScores[,2] ## select column 2
SecondColumn 
```



> ##Now you try! 
  - Make an object called "obj1" with a vector of the numbers 1-13
  - bind that vector to our "NamesScores" object, creating a new data frame
  - Name the new three column data frame "NamesScores2"
  
It might look something like this:
```{r, echo=FALSE}
obj1 <- c(1:13)
NamesScores2 <- cbind.data.frame(NamesScores, obj1)
NamesScores2
```

