---
title: "Data import and Tidy-ing"
author: Adapted from Data Science for R by Hadley Wichkam and Garrett Grolemund  
---


\



# Getting data into R! 

 *Maybe the most important step?*  
 
\  


##Base R 

* **read.csv()** is the function avaiable from base R, and maybe the one you are most familiar with  
    + This function reads data directly into a data frame so to use other packages (like dplyr) you will need to convert to a tibble yourself
    + You can specify if first columns of data are headers using header = TRUE  
```{r}
FAO1 <- read.csv("FAOSTAT_data.csv", header = TRUE)

```
\  



##readr
* **read_csv()** (from **readr** package) makes data from a .csv file right into a *tibble*  
    + up to 10x faster than **read.csv()**
    + reads in data directly as a tibble 

```{r }
library(tidyverse)

FAO1 <- read_csv("FAOSTAT_data.csv")
FAO1
```

#### Some tips for **read_csv()** 
  * Use skip = n to skip n nows if there are rows of metadata above your column headers. 
  * Use col_names = FALSE if there are no row headings or
  * col_names = c("x", "y", "z") to pass a vector of column names to the data table
  * use na = "." to specify what missing values look like  



Say that after import, you notice that you would like to change one of your variable types, you can do that by specifying the variable type for each column. For example, let's say we want to change year to character:
```{r}
FAO2 <- read_csv("FAOSTAT_data.csv", col_types = cols(
    Country = col_character(),
    Element = col_character(),
    Item = col_character(),
    Year = col_character(), ### changed year to character
    Unit = col_character(),
    Value = col_integer()
  ))
FAO2
```


### Read directly from excel files using the tidyverse package **readxl**  
* read_excel() detects file path 
*-* read_xlsx() reads directly from .xlsx file
```{r}
FAOxls <- readxl::read_excel("FAOSTAT_data2.xlsx")
FAOxls
```

### Read in other data types
- Read in SPSS, Stata and SAS files using package **haven**



### Parsing  
????   

\  
\  
\  
\  
\  
\  
????   









# Tidyr

- Tidy data has one variable per column 
- Each observation has a row
- Each value in it's own cell!   

#### In the book there are three data table examples. Which one is "tidy"?

```{r}
table1
table2
table3
```

#### Only table 1!  
#### Notice that each column is a variable. 

```{r}
table1
```


### Graph your tidy data  
####Using the data in Table 1, you can easily graph!   



```{r}
library(ggplot2)
ggplot(table1, aes(year, cases)) +
  geom_line(aes(group = country), color = "gray30") +
  geom_point(aes(color = country)) +
  theme_minimal()
```

#### Or, using our tidy FAO data, we can see how much acreage there was over time 

```{r}
FAO3 <- FAO2 %>% filter(Item == "Cassava")
library(ggplot2)
ggplot(FAO3, aes(Year, Value)) +
  geom_point(aes(color = Country)) +
  ylab("Area planted (ha)") +
  theme_minimal()
```

### Getting un-tidy data into shape


#### Spreading and gathering 
```{r}
table4a
table4a_2 <- table4a %>% gather ("1999", "2000", key = "year", value= "cases")
table4a_2
```

spread


#### Seperating and uniting

```{r}
MessyFAO <- read_csv("FAOSTAT_data.csv")
MessyFAO2 <- MessyFAO %>% unite(Area,Value,Unit, sep = "_") 
MessyFAO2

MessyFAO3 <- MessyFAO2 %>% separate(Area, into = c("Value", "Unit"))
MessyFAO3

```

